### 2016-07-05 introduction.tex
typos

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented
with a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
\emph{read committed} isolation level used in the experiments. The
same study has found that 13\% of all invariants among 67 open source
Ruby-on-Rails applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and
determine, in each case, if the phenomenon can be allowed without
violating application invariants. This is understandably hard,
especially in the absence of any formal underpinning to define weak
isolation semantics.  Adya~\cite{adyaphd} presents the first formal
definitions of some of well-known isolation levels in the context of a
sequentially consistent (SC) database.  However, there has been little
progress relating Adya's system model to a formal operational
semantics or a proof system that can facilitate rigorous correctness
arguments. Consequently, reasoning about weak isolation remains an
error prone endeavor, with major database vendors~\cite{postgresiso,
  mysqliso, oracleiso} continuing to document their isolation levels
primarily in terms of the undesirable phenomena a particular isolation
level may induce, placing the burden on the programmer to determine
application correctness.

@@

The remainder of the paper is organized as follows. The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:store-consistency} generalizes the
framework to integrate support for weakly consistent data stores. We
describe the impact of our reasoning framework in the context of
several real-world case studies in \S\ref{sec:case-studies}.  Related
work and conclusions are given in \S\ref{sec:relatedwork}.

>>>>

Unfortunately, weak isolation admits behaviours that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented
with a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
\emph{read committed} isolation level used in the experiments. The
same study has found that 13\% of all invariants among 67 open source
Ruby-on-Rails applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and
determine, in each case, if the phenomenon can be allowed without
violating application invariants. This is understandably hard,
especially in the absence of any formal underpinning to define weak
isolation semantics.  Adya~\cite{adyaphd} presents the first formal
definitions of some well-known isolation levels in the context of a
sequentially consistent (SC) database.  However, there has been little
progress relating Adya's system model to a formal operational
semantics or a proof system that can facilitate rigorous correctness
arguments. Consequently, reasoning about weak isolation remains an
error prone endeavour, with major database vendors~\cite{postgresiso,
  mysqliso, oracleiso} continuing to document their isolation levels
primarily in terms of the undesirable phenomena a particular isolation
level may induce, placing the burden on the programmer to determine
application correctness.

@@

The remainder of the paper is organized as follows. The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:store-consistency} generalizes the
framework to integrate support for weakly-consistent data stores. We
describe the impact of our reasoning framework in the context of
several real-world case studies in \S\ref{sec:case-studies}.  Related
work and conclusions are given in \S\ref{sec:relatedwork}.

### 2016-07-05 motivation.tex
Ids are not single quoted.
typos

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behavior of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under various levels of weak isolation,
determining what constitutes an interference is a non-trivial problem.
For example, a \iso{Serializable} transaction admits no interference,
whereas within a transaction executing under \iso{Read Uncommitted}
isolation, all interferences are valid. Between these two extremes are
various levels of isolation that admit some interferences while
prohibiting others. For example, \iso{Read Committed} isolation admits
interference of a committed transaction, but not those from
uncommitted transactions. \iso{Snapshot Isolation} admits interference
from committed non-conflicting transactions, and so on.

@@

We have thus far assumed a strongly consistent ({\sc sc}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behavior of {\sc sc} to totally
order all operations w.r.t. the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately visible to subsequent operations. This semantics can be
built into the reasoning framework, leading to a proof system
tailor-made for such stores. However, for the reasoning framework to
be truly useful, it has to be parametric over different consistency
semantics, supporting stores that are weaker than {\sc sc} (e.g.,
causally or eventually consistent), and should be able to reconcile
conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

>>>>

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behaviour of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under various levels of weak isolation,
determining what constitutes an interference is a non-trivial problem.
For example, a \iso{Serializable} transaction admits no interference,
whereas within a transaction executing under \iso{Read Uncommitted}
isolation, all interferences are valid. Between these two extremes are
various levels of isolation that admit some interferences while
prohibiting others. For example, \iso{Read Committed} isolation admits
interference of a committed transaction, but not those from
uncommitted transactions. \iso{Snapshot Isolation} admits interference
from committed non-conflicting transactions, and so on.

@@

We have thus far assumed a strongly consistent ({\sc sc}) store that,
in the absence of transactions with special isolation requirements,
makes the effects of any operation immediately visible to all
subsequent operations. The natural behaviour of {\sc sc} to totally
order all operations w.r.t. the $\hbZ$ relation could be in conflict
with the constraints imposed by weak isolation. To meet the
requirements of weaker isolation levels, say {\sc rc}, the store has
to adapt itself to \emph{hide} the effects of concurrent transactions
until they are committed. Once committed though, effects need to be
immediately visible to subsequent operations. This semantics can be
built into the reasoning framework, leading to a proof system
tailor-made for such stores. However, for the reasoning framework to
be truly useful, it has to be parametric over different consistency
semantics, supporting stores that are weaker than {\sc sc} (e.g.,
causally or eventually consistent), and should be able to reconcile
conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

### 2016-07-05 opsem.tex
typos

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behavior is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behavior as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:\vspace*{-8pt}

@@

As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behavior of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t. $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on application semantics. For example, a
monotonically increasing counter application may choose to let a write
with the largest value determine the value of a read. To accommodate
multiple interpretations, the semantics is made parametric over an
interpretation function ($\interp{\cdot}$) that accepts a set of
effects and a variable name, and returns the value associated with the
variable. A straightforward interpretation function that chooses the
last write (i.e., write with largest id) is shown below:

>>>>

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behaviour is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behaviour as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:\vspace*{-8pt}

@@

As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behaviour of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t. $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on application semantics. For example, a
monotonically increasing counter application may choose to let a write
with the largest value determine the value of a read. To accommodate
multiple interpretations, the semantics is made parametric over an
interpretation function ($\interp{\cdot}$) that accepts a set of
effects and a variable name, and returns the value associated with the
variable. A straightforward interpretation function that chooses the
last write (i.e., write with largest id) is shown below:

### 2016-07-05 store-consistency.tex
typos

The operational semantics of Fig.~\ref{fig:txnimp} allows operations
to witness arbitrary subsets of the global state, effectively mimicking the
behavior of an eventually consistent ({\sc ec}) data
store\footnote{Eventual consistency guarantees that in the absence of further
  updates, all reads witness the same global state
  \emph{eventually}. In any finite trace, however, there are no
  guarantees on what a read may witness.}. There are, however, data stores,
such as relational databases, that provide stronger consistency
guarantees than {\sc ec}. Like the isolation levels of transactions,
the consistency level of the underlying store also affects the
semantics of a program in non-trivial ways. In this section, we
demonstrate how the semantics of stronger stores with on-demand weak
isolation can be captured in our operational model.

>>>>

The operational semantics of Fig.~\ref{fig:txnimp} allows operations
to witness arbitrary subsets of the global state, effectively mimicking the
behaviour of an eventually consistent ({\sc ec}) data
store\footnote{Eventual consistency guarantees that in the absence of further
  updates, all reads witness the same global state
  \emph{eventually}. In any finite trace, however, there are no
  guarantees on what a read may witness.}. There are, however, data stores,
such as relational databases, that provide stronger consistency
guarantees than {\sc ec}. Like the isolation levels of transactions,
the consistency level of the underlying store also affects the
semantics of a program in non-trivial ways. In this section, we
demonstrate how the semantics of stronger stores with on-demand weak
isolation can be captured in our operational model.
